Why is it that statistical tests for residuals are not widely used? An explanation using visual inference.

### 2023-01-17 paper_comparison/paper_comparison/paper_comparison.tex
update main sections

One can access residuals in multiple ways. Residuals might be plotted,
as a histogram or quantile-quantile plot to examine the distribution.
Using the classical normal linear regression model as an example, if the
distribution is symmetric and unimodal, it is well-behaved. But if the
distribution is skewed, bimodal, multimodal, or contains outliers, there
is cause for concern. One could also inspect the distribution by
conducting a goodness of fit test, such as the Shapiro-Wilk Normality
test \citep{shapiro1965analysis}.

@@

\caption{Example fitted vs residual plots: (A) classically good looking residuals, (B) non-linear pattern indicates that the model has not captured a non-linear association, (C) heteroskedasticity indicating that variance around the fitted model is not uniform, and (D) non-normality where the residual distribution is not symmetric around 0. The latter pattern might best be assessed using a univariate plot of the residuals, but patterns B and C need to be assessed using a residual vs fitted plot.}\label{fig:residual-plot-common-departures}
\end{figure}

Graphical summaries in which residuals are plotted against fitted values
or other functions of the regressors that are approximately orthogonal
to residuals are referred to as standard residual plots in
\citet{cook1982residuals}. As shown in Figure
\ref{fig:residual-plot-common-departures}, the plot A is an ideal
residual plot with residuals evenly distributed at both sides of the
horizontal zero line, with no noticeable patterns.

@@

Non-linearity is a type of model misspecification caused by failing to
include higher order terms of the regressors in the regression equation.
Any non-linear functional form of residuals on fitted values presented
in the residual plot could be considered as an indicative of
non-linearity. An example residual plot containing visual pattern of
non-linearity is given at plot B of Figure
\ref{fig:residual-plot-common-departures}. One can clearly observe the
``S-shape'' from the residual plot as the cubic term is not captured by
the misspecified model.

Heteroskedasticity refers to the presence of nonconstant error variance
in a regression model. It is mostly due to the strict but false
assumptions on the variance-covariance matrix of the error term. The
usual pattern of heteroskedasticity on a residual plot is the
inconsistent spread of the residuals across the horizontal axis.
Visually, it sometimes results in the so-called ``butterfly'' shape as
shown in the plot C of Figure \ref{fig:residual-plot-common-departures},
or the ``left-triangle'' and ``right-triangle'' shape where the smallest
variance occurs at one side of the horizontal axis.

Compared to non-linearity and heteroskedasticity, non-normality is
usually harder to detect from a residual plot since a scatter plot do
not readily reveal the marginal distribution. A favourable graphical
summary for this task is the quantile-quantile plot. Hence,
non-normality will not be the focus of this paper as we mainly discuss
residual plots. For a consistent comparison, the residual plot of this
departure is still presented in plot D of Figure
\ref{fig:residual-plot-common-departures}. Besides, it is important to
note that not all regression models assume normality for the error term,
but a certain amount do including the classical normal linear regression
model. In the case that the normality assumption is violated, it is
expected to observe data points do not center around the horizontal axis
and there is an uneven distribution of the number points at both below
and above the horizontal axis. For example, given a skewed error
distribution, there will be fewer data points and more outliers on one
side of the horizontal axis as shown in plot D of Figure
\ref{fig:residual-plot-common-departures}.

@@

\caption{\label{tab:example-residual-plot-table}Statistical significance testing for departures from good residuals for plots in Figure \ref{fig:residual-plot-common-departures}. Shown are the $p$-values calculated for the conventional RESET, the Breusch-Pagan and the Shapiro–Wilk tests. The good residual plot (A) is judged a good residual plot, as expected, by all tests. The non-linearity (B) is detected by all tests, as might be expected given the extreme structure.}
\centering
\begin{tabular}[t]{llrrr}
\toprule
Plot & Departures & RESET & Breusch-Pagan & Shapiro–Wilk\\
\midrule
A & None & 0.779 & 0.133 & 0.728\\
B & Non-linearity & \em{0.000} & \em{0.000} & \em{0.039}\\
C & Heteroskedasticity & 0.658 & \em{0.000} & \em{0.000}\\
D & Non-normality & 0.863 & 0.736 & \em{0.000}\\
\bottomrule
\end{tabular}
\end{table}

@@

Visual inference was first introduced in a 1999 Joint Statistical
Meetings (JSM) talk with the title ``Inference for Data Visualization''
by \citet{buja_inference_1999} as an idea to address the issue of valid
inference for visual discoveries of data plots. Later,
\citet{buja_statistical_2009} proposed the lineup protocol as a visual
test inspired by the ``police lineup'' or ``identity parade'' which is
the act of asking the eyewitness to identify criminal suspect from a
group of irrelevant people. The protocol consists of \(m\) randomly
placed plots, where one plot is the data plot, and the remaining
\(m - 1\) plots have the identical graphical procedure except the data
has been replaced with data consistent with \(H_0\). Then, an observer
who have not seen the data plot will be asked to point out the most
different plot from the lineup. Under \(H_0\), it is expected that the
data plot would have no distinguishable difference from the null plots,
and the probability that the observer correctly picks the data plot is
\(1/m\). If one rejects \(H_0\) as the observer correctly picks the data
plot, then the Type I error of this test is \(1/m\).

@@

The effectiveness of lineup protocol for regression analysis is
validated by \citet{majumder_validation_2013} under relatively simple
settings with up to two regressors. Their results suggest that visual
tests are capable of testing the significance of a single regressor with
a similar power as a t-test, though they express that in general it is
unnecessary to use visual inference if there exists a conventional test,
and they do not expect the visual test to perform equally well as the
conventional test. In their third experiment, where there is not a
conventional test, visual test outperforms the conventional test for a
large margin. This is encouraging, as it promotes the use of visual
inference in situations where there are no existing statistical testing
procedures. Visual inference have also been integrated into diagnostic
of hierarchical linear models by \citet{loy2013diagnostic},
\citet{loy2014hlmdiag} and \citet{loy2015you}. They use lineup protocols
to judge the assumption of linearity, normality and constant error
variance for both the level-1 and level-2 residuals. (expand?)

@@

Data used in the \(m - 1\) null plots needs to be simulated. In the
context of regression diagnostics, sampling data from \(H_0\) is
equivalent to sampling data from the assumed model. As
\citet{buja_statistical_2009} suggested, \(H_0\) is usually a composite
hypothesis controlled by nuisance parameters. Since regression models
can have various forms, there is no general solution to this problem,
but it sometimes can be reduced to a so called ``reference
distribution'' by applying one of the three methods: (i) sampling from a
conditional distribution given a minimal sufficient statistic under
\(H_0\), (ii) parametric bootstrap sampling with nuisance parameters
estimated under \(H_0\), and (iii) Bayesian posterior predictive
sampling. The conditional distribution given a minimal sufficient
statistic is the best justified reference distribution among the three
\citep{buja_statistical_2009}. Essentially, null residuals can be
simulated by regressing \(N\) i.i.d standard normal random draws on the
regressors, then rescaling it by the ratio of residual sum of square in
two regressions.

@@

In hypothesis testing, \(p\)-value is defined as the probability of
observing test results as least as extreme as the observed result given
\(H_0\) is true. Within the context of visual inference, by involving
\(k\) independent observers, \(p\)-value can be interpreted as the
probability of having as many or more subjects detect the data plot than
the observed result.

Let \(X_j = \{0,1\}\) be a binomial random variable denoting whether
subject \(j\) correctly detecting the data plot, and
\(X = \sum_{j=1}^{K}X_j\) be the number of observers correctly picking
the data plot. Then, by imposing a relatively strong assumption on the
visual test that all \(K\) evaluations are fully independent, under
\(H_0\), \(X \sim \mathrm{Binom}_{K,1/m}\). Therefore, the \(p\)-value
of a lineup of size \(m\) evaluated by \(K\) observer is given as
\(P(X \geq x) = 1 - F(x) + f(x)\), where \(F(.)\) is the cumulative
distribution function, \(f(.)\) is the probability mass function and
\(x\) is the realization of number of observers correctly picking the
data plot \citep{majumder_validation_2013}.

@@

The power of a model misspecification test specifies the chance of a
misspecified regression model being detected given a certain alternative
hypothesis. It is an important indicator when one concerns if the
regression model is misspecified. Though, people might usually be more
interested in knowing how much the residuals depart from the model
assumptions and if the departure is of practical significance.

As discussed in \citet{majumder_validation_2013}, individual's skill may
affect the number of observers who identify the data plot from the
lineup. Thus, the power of a visual test could depend on the
subject-specific abilities. Previously, it is addressed by modelling the
probability of a subject \(j\) correctly picking the data plot from a
lineup \(l\) using a mixed-effect logistic regression with the subject
being treated as a random effect \citep{majumder_validation_2013}.
However, in the multiple selections scenario, having this probability is
insufficient to determine the power of a visual test as it doesn't
provide information about the number of selections made by the subject
for p-value calculation.

Instead, we estimate the probability of a lineup being rejected directly
by assuming the individual skill has negligible impact on the variation
of the power of a lineup. The assumption is made to simplify the model
structure, thus mitigating the need for expensive large-scale
experiments to estimate the complex covariance matrix. The model is a
logistic regression with the natural logarithm of the effect as the only
regressor formulated as

\begin{equation} \label{eq:logistic-regression-1-1}
Pr(\text{reject}~H_0|H_1,E) = \Lambda(\beta_0 + \beta_1 log_e(E)),
\end{equation}

@@

\begin{equation} \label{eq:effect-size-ex1}
E = \frac{1}{2\sigma^2}\boldsymbol{X}_b'\boldsymbol{R}_a'(diag(\boldsymbol{R}_a))^{-1}\boldsymbol{R}_a\boldsymbol{X}_b,
\end{equation}

\noindent where \(diag(.)\) is the diagonal matrix constructed from the
diagonal elements of \(\boldsymbol{R}_a\).

To study various factors contributing to the power of the visual test,
the same logistic regression model is fit on different subsets of the
collated data grouped by levels of factors. This includes
{[}expansion{]}.

@@

\begin{align} \label{eq:heter-model}
\boldsymbol{y} = 1 + \boldsymbol{x} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} = g(\boldsymbol{x}_{raw}, 1)\\
\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}, 1 + (2 - |a|)(\boldsymbol{x} - a)^2b \boldsymbol{I}), \\
\end{align}

@@

@@

Figure \ref{fig:polypower} shows the estimated power of visual test on
lineups produced by the non-linearity model with
\(X_{raw} \sim U(-1,1)\), against the natural logarithm of the effect
\(log_e(E)\), with a 5\% significance level. Uniform distribution is the
focus of this section because visual patterns are more likely to be
revealed. Besides, six more evaluations are collected for the uniform
distribution in experiment III, which will produce more reliable and
stable power curves. Analysis of other distributions can be found in
Section \ref{effect-of-the-distribution-of-the-regressor}. At the bottom
of the figure \ref{fig:polypower}, there are a sequence of example
residual plots with increasing levels of \(log_e(E)\). Readers can
evaluate them from left to right and determine at which level the
departure from a good residual plot becomes detectable.

@@

For the non-linearity model, the power curve of RESET test climbs
aggressively from 13\% to around 70\% as \(log_e(E)\) increases from 0
to 2, while power of other tests respond inactively to the change of
effect, showing that REST test is way more sensitive to the type of
model defects that being considered. Meanwhile, no noticeable visual
features can be spotted from the example residual plots.

@@

Depending on the natural logarithm of the effect, lineups are classified
into three categories: easy (\textgreater{} 3.5), moderate (1.5 - 3.5)
and difficult (\textless{} 1.5). Figure \ref{fig:p-value-comparison}
provides a mosaic plot showing the rejection rate of visual tests and
conventional tests at different difficulty levels.

\hypertarget{easy-lineup}{%
\subsubsection{Easy lineup}\label{easy-lineup}}

For easy lineups produced by the non-linearity model, 65\% are rejected
by both the conventional and visual tests, while 9\% are accepted by
both tests. Thus, 74\% of the time two tests will agree with each other.
There are 16\% of lineups rejected by conventional tests only.

Interestingly, in spite of the greater power of conventional tests, 10\%
of the lineups are rejected by visual tests only. Recall that the RESET
test used in our analysis includes powers of fitted values up to four.
However, the ``M'' and the ``Triple-U'' shapes constructed from the
Hermite polynomials contain power of the regressor up to six and 18
respectively. In fact, by including the power of fitted values up to six
in the RESET test, all these cases will be rejected. We also can observe
from the example plot displayed in Figure \ref{fig:easy-poly-example}
that panel three is distinctly different from others, exhibiting a ``M''
shape. This clearly suggests users of the RESET test and any other
residual-based conventional tests that require the specification of
variables of interest to check the residual plot before conducting the
test, such that the correct choice of variables can be made.

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/easy-poly-example-1}

\caption{Lineup poly-270 with $log_e(E) = 3.89$ produced by the non-linearity model used in experiment I and III. It is rejected by the visual test but not by the RESET test. The data plot at panel three exhibits strong non-linear pattern.}\label{fig:easy-poly-example}
\end{figure}

For easy lineups produced by the heteroskedasticity model, 62\% are
rejected by both the conventional test and the visual test, and 7\% are
accepted by both, result in 69\% agreement rate. Lineups only rejected
by conventional test are 30\%. 1\% of lineups are accepted by the
conventional test but rejected by the visual test.

Among those 1\%, which are two lineups, one shows detectable
heteroskedasticity pattern as displayed in Figure
\ref{fig:easy-heter-example1}. Another contain data plot with an
noticeable outlier, but the ``Right-triangle'' shape is not very strong
as shown in Figure \ref{fig:easy-heter-example2}. The \(p\)-values of
Breusch-Pagan test for these two lineups are \(0.055\) and \(0.135\).
Compared to the visual test, its \(p\)-values are \(0.026\) and
\(0.012\). The Breusch-Pagan test should reject the one with obvious
heteroskedasticity signal, but it yields a \(p\)-value slightly above
5\%.

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/easy-heter-example1-1}

\caption{Lineup heter-331 with $log_e(E) = 4.02$ produced by the heteroskedasticity model used in experiment II and III. It is rejected by the visual test but not by the  Breusch-Pagan test. The data plot at panel 17 contains a "Butterfly" shape, indicating the presence of heteroskedasticity.}\label{fig:easy-heter-example1}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/easy-heter-example2-1}

\caption{Lineup heter-536 with $log_e(E) = 4.13$ produced by the heteroskedasticity model used in experiment II. It is rejected by the visual test but not by the  Breusch-Pagan test. The data plot at panel 18 exhibits a vague heteroskedasticity pattern, but it contains a distinct outlier which makes it attractive to subjects.}\label{fig:easy-heter-example2}
\end{figure}

\hypertarget{moderate-lineup}{%
\subsubsection{Moderate lineup}\label{moderate-lineup}}

With the increase of the difficulty level, the sensitivity of
conventional tests enlarge the difference in rejection rates. For
moderate lineup produced by the non-linearity model, 8\% and 28\% are
rejected and accepted by both tests respectively. We can observe a
decline in agreement rate to 36\%. A large proportion of lineups are
accepted by the visual test and rejected by the conventional test, which
is around 63\%. The remaining 1\% are rejected by conventional tests
only, which suffer the same issue caused by the RESET test as mentioned
earlier.

For those produced by the heteroskedasticy model, the corresponding
numbers are 23\%, 31\%, 54\%, 45\% and 1\%. There is only one lineup
rejected by the visual test but accepted by the conventional test. It
does not show any heteroskedasticity patterns, but contains a few
outliers at the bottom of the data plot as illustrated in panel 14 of
Figure \ref{fig:moderate-heter-example1}. Two out of five subjects
correctly identify this lineup yielding a \(p\)-value equal to
\(0.044\). We consider this instance is rejected by chance. If a
different set of null plots are used, the data plot may not be
identified.

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/moderate-heter-example1-1}

\caption{Lineup heter-219 with $log_e(E) = 1.61$ produced by the heteroskedasticity model used in experiment II. It is rejected by the visual test but not by the  Breusch-Pagan test. Though the data plot at panel 14 shows no heteroskedasticity patterns, it contains a few outliers at the bottom of the panel.}\label{fig:moderate-heter-example1}
\end{figure}

\hypertarget{difficult-lineup}{%
\subsubsection{Difficult lineup}\label{difficult-lineup}}

Visual test is insensitive to lineups with small effect. When it comes
to difficult level, none of the lineups produced by the non-linearity
model are rejected by the visual test. There are 73\% of lineups
accepted by both tests, and 26\% rejected by the conventional test. The
agreement rate rise up to 73\%.

For lineups produced by the heteroskedasticity model, 3\% are rejected
by both tests, and 71\% are accepted by both tests. There are 24\% of
lineups rejected by the conventional test only. Besides, only two
lineups, which is 2\% of all the difficult lineups, are rejected by the
visual test only.

As shown in \ref{fig:difficult-heter-example1} and
\ref{fig:difficult-heter-example2}, residual plots of both lineups are
constructed with \(X_{raw} \sim lognormal(0, 0.6^2)/3\). Though the
designed patterns are ``Left-triangle'' and ``Butterfly'' shape, both of
them are invisible due to the small effect. The visual effect are
dominated by the skewed distribution. Further, no distinguishable visual
patterns, such as outliers, are detectable from the data plot of these
two lineups, which are panel 5 and panel 15. The data plots in these two
lineups are not the most different plot from our point of view either.
However, three out of five and two out of five subjects correctly
identify the data plot for these two lineups respectively. But the
reasons for selecting the plot provided by the subjects are mixed and
inconsistent. Hence, with the current data, we can not explain why
subjects can correctly identify the data plot of these two lineups.
Collecting more data in future study may help to answer this question.

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/difficult-heter-example1-1}

\caption{Lineup heter-224 with $log_e(E) = 1.32$ produced by the heteroskedasticity model used in experiment II. It is rejected by the visual test but not by the  Breusch-Pagan test. Although two out of five subjects correctly identify the data plot at panel 5, it does not exhibit any visually distinguishable pattern.}\label{fig:difficult-heter-example1}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/difficult-heter-example2-1}

\caption{Lineup heter-415 with $log_e(E) = -0.31$ produced by the heteroskedasticity model used in experiment II. It is rejected by the visual test but not by the  Breusch-Pagan test. Panel 15 is the data plot but it does not exhibit any visually distinguishable pattern.}\label{fig:difficult-heter-example2}
\end{figure}

@@

\caption{Rejection rate of visual test compared to conventional test on lineups produced by heteroskedasticity and non-linearity model at different levels of difficulty. Based on the rejection decision made by the visual test and the conventional test, the mosaic plot shows the proportion of lineups in each category.}\label{fig:p-value-comparison}
\end{figure}

@@

@@

Figure \ref{fig:different-x-dist-poly-power} and Figure
\ref{fig:different-x-dist-heter-power} illustrate the change of power of
visual tests and conventional tests for different distribution of
\(X_{raw}\) used in the non-linearity model and the heteroskedasticity
model.

>>>>

We can assess residuals in multiple ways. Residuals might be plotted, as
a histogram or quantile-quantile plot to examine the distribution. Using
the classical normal linear regression model as an example, if the
distribution is symmetric and unimodal, we consider it well-behaved. But
if the distribution is skewed, bimodal, multimodal, or contains
outliers, there is cause for concern. One could also inspect the
distribution by conducting a goodness of fit test, such as the
Shapiro-Wilk Normality test \citep{shapiro1965analysis}.

@@

\caption{Example residual vs fitted plots: (A) classically good looking residuals, (B) non-linear pattern indicates that the model has not captured a non-linear association, (C) heteroskedasticity indicating that variance around the fitted model is not uniform, and (D) non-normality where the residual distribution is not symmetric around 0. The latter pattern might best be assessed using a univariate plot of the residuals, but patterns B and C need to be assessed using a residual vs fitted plot.}\label{fig:residual-plot-common-departures}
\end{figure}

Graphical summaries in which residuals are plotted against fitted values
or other functions of the regressors that are approximately orthogonal
to residuals are referred to as standard residual plots in
\citet{cook1982residuals}. Figure
\ref{fig:residual-plot-common-departures}A shows an ideal residual plot
with residuals evenly distributed at both sides of the horizontal zero
line, with no noticeable patterns.

@@

Non-linearity is a type of model misspecification caused by failing to
include higher order terms of the regressors in the regression equation.
Any non-linear functional form of residuals on fitted values in the
residual plot could be indicative of non-linearity. An example residual
plot containing visual pattern of non-linearity is given at Figure
\ref{fig:residual-plot-common-departures}B. One can clearly observe the
``S-shape'' from the residual plot as the cubic term is not captured by
the misspecified model.

Heteroskedasticity refers to the presence of nonconstant error variance
in a regression model. It is mostly due to the strict but false
assumptions on the variance-covariance matrix of the error term. The
usual pattern of heteroskedasticity on a residual plot is the
inconsistent spread of the residuals across the horizontal axis.
Visually, it sometimes results in the so-called ``butterfly'' shape as
shown in the Figure \ref{fig:residual-plot-common-departures}C, or the
``left-triangle'' and ``right-triangle'' shape where the smallest
variance occurs at one side of the horizontal axis.

Compared to non-linearity and heteroskedasticity, non-normality is
usually harder to detect from a residual plot since a scatter plot do
not readily reveal the marginal distribution. A favourable graphical
summary for this task is the quantile-quantile plot. As we mainly
discuss residual plots, non-normality will not be the focus of this
paper. For a consistent comparison, the residual plot of this departure
is still presented in Figure \ref{fig:residual-plot-common-departures}D.
When the number residuals below and above the horizontal axis are uneven
across the local regions along the \(x\)-axis, we expect that the
normality assumption is violated. For example, given a skewed error
distribution, there will be fewer data points and more outliers on one
side of the horizontal axis as shown in Figure
\ref{fig:residual-plot-common-departures}D.

@@

\caption{\label{tab:example-residual-plot-table}Statistical significance testing for departures from good residuals for plots in Figure \ref{fig:residual-plot-common-departures}. Shown are the $p$-values calculated for the RESET, the Breusch-Pagan and the Shapiro–Wilk tests. The good residual plot (A) is judged a good residual plot, as expected, by all tests. The non-linearity (B) is detected by all tests, as might be expected given the extreme structure.}
\centering
\begin{tabular}[t]{llrrr}
\toprule
Plot & Departures & RESET & Breusch-Pagan & Shapiro–Wilk\\
\midrule
A & None & 0.779 & 0.133 & 0.728\\
B & Non-linearity & \em{0.000} & \em{0.000} & \em{0.039}\\
C & Heteroskedasticity & 0.658 & \em{0.000} & \em{0.000}\\
D & Non-normality & 0.863 & 0.736 & \em{0.000}\\
\bottomrule
\end{tabular}
\end{table}

@@

Visual inference was first introduced in a 1999 Joint Statistical
Meetings (JSM) talk with the title ``Inference for Data Visualization''
by \citet{buja_inference_1999} as an idea to address the issue of valid
inference for visual discoveries of data plots. Later,
\citet{buja_statistical_2009} proposed the lineup protocol as a visual
test inspired by the ``police lineup'' or ``identity parade'' which is
the act of asking the eyewitness to identify criminal suspect from a
group of irrelevant people. The protocol consists of \(m\) randomly
placed plots, where one plot is the data plot, and the remaining
\(m - 1\) null plots have the identical graphical procedure except the
data has been replaced with data consistent with \(H_0\). Then, an
observer who have not seen the data plot will be asked to point out the
most different plot from the lineup. Under \(H_0\), it is expected that
the data plot would have no distinguishable difference from the null
plots, and the probability that the observer correctly picks the data
plot is \(1/m\). If one rejects \(H_0\) as the observer correctly picks
the data plot, then the Type I error of this test is \(1/m\).

@@

The effectiveness of lineup protocol for regression analysis is
validated by \citet{majumder_validation_2013} under relatively simple
settings with up to two regressors. Their results suggest that visual
tests are capable of testing the significance of a single regressor with
a similar power as a t-test, though they express that in general it is
unnecessary to use visual inference if there exists a conventional test,
and they do not expect the visual test to perform equally well as the
conventional test. (contradict to our claim? expand?) In their third
experiment, where there is not a conventional test, visual test
outperforms the conventional test for a large margin. This is
encouraging, as it promotes the use of visual inference in situations
where there are no existing statistical testing procedures. Visual
inference have also been integrated into diagnostic of hierarchical
linear models by \citet{loy2013diagnostic}, \citet{loy2014hlmdiag} and
\citet{loy2015you}. They use lineup protocols to judge the assumption of
linearity, normality and constant error variance for both the level-1
and level-2 residuals. (expand?)

@@

Data used in the \(m - 1\) null plots needs to be simulated. In
regression diagnostics, sampling data consistent with \(H_0\) is
equivalent to sampling data from the assumed model. As
\citet{buja_statistical_2009} suggested, \(H_0\) is usually a composite
hypothesis controlled by nuisance parameters. Since regression models
can have various forms, there is no general solution to this problem,
but it sometimes can be reduced to a so called ``reference
distribution'' by applying one of the three methods: (i) sampling from a
conditional distribution given a minimal sufficient statistic under
\(H_0\), (ii) parametric bootstrap sampling with nuisance parameters
estimated under \(H_0\), and (iii) Bayesian posterior predictive
sampling. The conditional distribution given a minimal sufficient
statistic is the best justified reference distribution among the three
\citep{buja_statistical_2009}. Essentially, null residuals can be
simulated by regressing \(N\) i.i.d standard normal random draws on the
regressors, then rescaling it by the ratio of residual sum of square in
two regressions.

@@

In hypothesis testing, a \(p\)-value is defined as the probability of
observing test results as least as extreme as the observed result given
\(H_0\) is true. Within the context of visual inference, by involving
\(k\) independent observers, the \(p\)-value can be interpreted as the
probability of having as many or more subjects detect the data plot than
the observed result.

Let \(X_j = \{0,1\}\) be a Bernoulli random variable denoting whether
subject \(j\) correctly detecting the data plot, and
\(X = \sum_{j=1}^{K}X_j\) be the number of observers correctly picking
the data plot. Then, by imposing a relatively strong assumption on the
visual test that all \(K\) evaluations are fully independent, under
\(H_0\), \(X \sim \mathrm{Binom}_{K,1/m}\). Therefore, the \(p\)-value
of a lineup of size \(m\) evaluated by \(K\) observer is given as
\(P(X \geq x) = 1 - F(x) + f(x)\), where \(F(.)\) is the cumulative
distribution function, \(f(.)\) is the probability mass function and
\(x\) is the realization of number of observers correctly picking the
data plot \citep{majumder_validation_2013}.

@@

The power of a model misspecification test is the probability that the
null hypothesis is rejected given the regression model is misspecified.
It is an important indicator when one is concerned about whether model
assumptions have been violated. Although in practice, one might be more
interested in knowing how much the residuals deviate from the model
assumptions, and whether this deviation is of practical significance.

As discussed in \citet{majumder_validation_2013}, the power of a visual
test may depend on the ability of the particular subject, as the skill
of the individual may affect the number of observers who identify the
data plot from the lineup. Previously, it is addressed by modelling the
probability of a subject \(j\) correctly picking the data plot from a
lineup \(l\) using a mixed-effect logistic regression, with subjects
treated as random effects \citep{majumder_validation_2013}. However, in
the multiple selections scenario, having this probability is not
sufficient to determine the power of a visual test because it does not
provide information about the number of selections made by the subject
for the calculation of the p-value.

Instead, we directly estimate the probability of a lineup being rejected
by assuming that individual skill has negligible effect on the variation
of the power. The assumption is to simplify the model structure, thereby
obviate costly large-scale experiments to estimate complex covariance
matrices. The model is a logistic regression with the natural logarithm
of the effect as the only regressor formulated as

\begin{equation} \label{eq:logistic-regression-1-1}
Pr(\text{reject}~H_0|H_1,E) = \Lambda(\beta_0 + \beta_1 log(E)),
\end{equation}

@@

\small

\begin{gather}  \label{eq:effect-size}
E = \frac{1}{2}\left(log\frac{|diag(\boldsymbol{R}_a\boldsymbol{V})|}{|diag(\boldsymbol{R}_a\sigma^2)|} - n + tr(diag(\boldsymbol{R}_a\boldsymbol{V})^{-1}diag(\boldsymbol{R}_a\sigma^2)) + \boldsymbol{\mu}_z^{T}(\boldsymbol{R}_a\boldsymbol{V})^{-1}\boldsymbol{\mu}_z\right),\\
\boldsymbol{R}_a = \boldsymbol{I}_n - \boldsymbol{X}_a(\boldsymbol{X}_a^T\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a^T,
\end{gather}

\normalsize

\noindent where \(diag(.)\) is the diagonal matrix constructed from the
diagonal elements of a matrix,
\(\boldsymbol{X}_a = (\boldsymbol{1}, \boldsymbol{X})\) is the matrix of
regressors including the intercept used in the regression equation,
\(\sigma^2\) is the assumed variance of the error term when \(H_0\) is
true, \(\boldsymbol{V}\) is the actual variance of the error term, and
\(\boldsymbol{\mu}_z = \boldsymbol{R}_a\boldsymbol{Z}\boldsymbol{\beta}_z\)
is the expected values of the residuals with \(\boldsymbol{Z}\) be any
variables leave out by the model and \(\boldsymbol{\beta}_z\) be the
corresponding coefficients.

To study various factors contributing to the power of the visual test,
the same logistic regression model is fit on different subsets of the
collated data grouped by levels of factors. These include the
distribution of the regressor and the type of the simulated model.

@@

\begin{align} \label{eq:heter-model}
\boldsymbol{y} &= 1 + \boldsymbol{x} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} &= g(\boldsymbol{x}_{raw}, 1),\\
\boldsymbol{\varepsilon} &\sim N(\boldsymbol{0}, 1 + (2 - |a|)(\boldsymbol{x} - a)^2b \boldsymbol{I}), 
\end{align}

@@

In the following analysis, lineups with \(X_{raw} \sim U(-1,1)\) will be
the focus of next two sections, because visual patterns are more likely
to be revealed with uniform distribution. Besides, six more evaluations
are collected for the uniform distribution in experiment III, which will
produce more reliable and stable results. Analysis of other
distributions can be found in Section
\ref{effect-of-the-distribution-of-the-regressor}.

@@

Figure \ref{fig:polypower} shows the estimated power of visual test on
lineups produced by the non-linearity model with
\(X_{raw} \sim U(-1,1)\), against the natural logarithm of the effect
\(log_e(E)\), with a 5\% significance level. At the bottom of the figure
\ref{fig:polypower}, there are a sequence of example residual plots with
increasing levels of \(log_e(E)\). Readers can evaluate them from left
to right and determine at which level the departure from a good residual
plot becomes detectable.

@@

For the non-linearity model, the power curve of RESET test climbs
aggressively from 13\% to around 70\% as \(log_e(E)\) increases from 0
to 2, while power of other tests respond inactively to the change of
effect, showing that RESET test is way more sensitive to the type of
model defects that being considered. Meanwhile, no noticeable visual
features can be spotted from the example residual plots.

@@

@@

\caption{Rejection rate of visual test compared to conventional test on lineups produced by heteroskedasticity and non-linearity model. Based on the rejection decision made by the visual test and the conventional test, the mosaic plot shows the proportion of lineups in each category.}\label{fig:p-value-comparison}
\end{figure}

(remove lineup id from caption)

Figure \ref{fig:p-value-comparison} provides a mosaic plot showing the
rejection rate of visual tests and conventional tests for both the
non-linearity model and the heteroskedasticity model.

For lineups produced by the non-linearity model, 25\% are rejected by
both the conventional and visual tests, while 42\% are accepted by both
tests. Thus, 69\% of the time two tests will agree with each other.
There are 26\% of lineups rejected by conventional tests only.

Interestingly, in spite of the greater power of conventional tests, 7\%
of the lineups are rejected by visual tests only. Recall that the RESET
test used in our analysis includes powers of fitted values up to four.
However, the ``M'' and the ``Triple-U'' shapes constructed from the
Hermite polynomials contain power of the regressor up to six and 18
respectively. In fact, by including the power of fitted values up to six
in the RESET test, all these cases will be rejected. We also can observe
from the example plot displayed in Figure \ref{fig:poly-example} that
panel three is distinctly different from others, exhibiting a ``M''
shape. This clearly suggests users of the RESET test and any other
residual-based conventional tests that require the specification of
variables of interest to check the residual plot before conducting the
test, such that the correct choice of variables can be made.

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/poly-example-1}

\caption{Lineup poly-270 with $log_e(E) = 3.89$ produced by the non-linearity model used in experiment I and III. It is rejected by the visual test but not by the RESET test. The data plot at panel three exhibits strong non-linear pattern.}\label{fig:poly-example}
\end{figure}

\begin{center}\includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/unnamed-chunk-11-1} \end{center}

\begin{center}\includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/unnamed-chunk-12-1} \end{center}

\begin{center}\includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/unnamed-chunk-13-1} \end{center}

For lineups produced by the heteroskedasticity model, 55\% are rejected
by both the conventional test and the visual test, and 24\% are accepted
by both, result in 79\% agreement rate. Percentage of lineups only
rejected by conventional test are 20\%. Besides, only 1 lineup, which is
less than 1\% of all the lineups, are rejected by the visual test only.
This lineup shows relatively strong heteroskedasticity pattern as
displayed in Figure \ref{fig:heter-example}. The visual test rejects it
with a \(p\text{-value} = 0.026\), but the Breusch-Pagan test produces a
\(p\)-value slightly above 5\%. However, as we can see, it is an
exceptional case.

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/heter-example-1}

\caption{Lineup heter-331 with $log_e(E) = 4.02$ produced by the heteroskedasticity model used in experiment II and III. It is rejected by the visual test but not by the  Breusch-Pagan test. The data plot at panel 17 contains a "Butterfly" shape, indicating the presence of heteroskedasticity.}\label{fig:heter-example}
\end{figure}

@@

(calc the diff in power and plot)

(randomly subset uniform and check the change of the power)

@@

Figure \ref{fig:different-x-dist-poly-power} and Figure
\ref{fig:different-x-dist-heter-power} illustrate the change of power of
visual tests and conventional tests for different distribution of
\(X_{raw}\) used in the non-linearity model and the heteroskedasticity
model. We focus on \(log(E) > 2\) as the visual patterns become visible.

For the non-linearity model, we do not observe significant difference of
power of visual test between distributions, except for the discrete
uniform distribution. It could be due to the fact that humans have
difficulties in recognising visual patterns from discreteness, which
makes the shape disconnected and incomplete (ref?). In terms of
conventional tests, the discrete uniform distribution is instead the one
with the greatest power, followed by the lognormal distribution, uniform
distribution and the normal distribution.

For the heteroskedasticity model, the power of visual test on lineups
with uniform distribution has the greatest power. This is as expected
since other three distributions could reduce the chance of revealing the
underlying visual pattern because of uneven data points. Although the
power of visual test under the discrete uniform distribution is the
lowest in the case of non-linearity model, it has a relatively great
power this time. Considering for heteroskedasticity, the visual patterns
are usually detected by connecting the maximum and minimum residuals
separately at different fitted values. It will not be greatly affected
by the discrete uniform distribution compared to the uniform
distribution. Visual tests have low power on lineups with the normal
distribution and the lognormal distribution. This suggests the use of
transformation of fitted values in reading residual plots. In the case
of conventional tests, the discrete uniform distribution consistently
doing great, followed by the uniform distribution, the normal
distribution and the lognormal distribution.
